Operational Feature Selection in Gaussian Mixture Models

### 2016-09-28 paper.tex
Intro FS

    The computation of~\ref{eq:decision} requires the inversion of the
    covariance matrix and and the  computation of the logarithm of the
    determinant.  The estimations of these terms suffer from the curse
    of  dimensionality\cite{}, because  in practice,
    the  number of  parameters  $\eta_c$ to  estimate  for each  class
    increases quadratically with respect to the number of features, as
    illustrated in Figure~\ref{fig:nb-param}.  Hence, if the number of
    observation  $n_c$ is  small  compared to  the  number of  samples
    $n_c$, the  estimated covariance  matrix is badly  conditioned and
    thus the computation  of its inverse and its  determinant would be
    unstable.   The worst  situation  is $n_c<\eta_c$  which leads  to
    singular covariance matrix.  Unfortunately, this situation happens
    regularly in remote sensing.   For instance in hyperspectral image
    classification,  very few  labeled samples  are usually  available
    because of the difficulty and the cost to collect ground-truth.

@@

Features selection algorithms may be divided into three types. The first type called filter method is based uniquely on data analysis. Features are ranked according to a statistical analysis of the data. For example, the Principal Component Analysis (PCA) described in \cite{jimenez1998supervised} is a typical filter method and there is numerous other methods \cite{bruzzone1995extension}, \cite{biesiada2007feature}, \cite{demir2008phase}.

The second sort are known as wrapper methods which can be seen as search method to determine the best subset of variables for a given learning model. As exhaustive searches are out of question in a practical amount of time, numerous search strategies have been designed some optimal under particular hypothesis \cite{narendra1977branch} and other suboptimal but easier to set up \cite{whitney1971direct}, \cite{somol1999adaptive}. The advantage of such methods compare to filter methods is that they are dedicated to a particular model but on the other hand, as they require the training of multiple models to test various set of variables, they tend to be slower.

The third type corresponds to the embedded method which do not separate the features selection process from the learning algorithm and allow interaction between the two processes. The basic example of such method is the decision tree algorithm in which a feature is selected for the creation of each node. Embedded methods also exist for other models, e.g. SVM \cite{guyon2002gene} \cite{weston2003use}.

The selection method proposed in this work is a wrapper method associated to GMM models. Thus, in order to set a wrapper method, two elements are needed, a function to rank the various features which is defined in Section~\ref{sec:criterion} and a search strategy. Two search algorithms are presented: the sequential forward features selection method (Section \ref{sec:forward-presentation}) and the sequential floating forward feature selection method (\ref{sec:floating-presentation}) the second being a more complex variation of the first.

@@

    \subsection{Selection method}

@@

\section{Datasets}
\label{sec:datasets}
Numerical experiments  have been conducted on  two different datasets.
The first  dataset, called  \emph{Aisa}, is an  airborne hyperspectral
dataset and  the second,  called \emph{Potsdam},  is an  airborne very
high resolution multispectral image.

@@

    \begin{table}[!t]
        \centering
        \caption{Kappa coefficient and processing time for 1,000 samples by class and averaged over 5 trials (standard deviation in parenthesis). Processing times are given in second.\label{tab:potsdam-otbsimu}}
        \begin{tabular}{lcccccc}\toprule
            & {\bfseries 5\_11 (train)} & {\bfseries 5\_12 (test)} & {\bfseries 3\_10 (test)} & {\bfseries Train. Time} & {\bfseries Classif. time} & {\bfseries \# of selected features} \\ \cline{2-6}
            GMM SFS kappa & 0.694 (0.002)             & {\bfseries 0.669 (0.005)} & {\bfseries 0.533 (0.008)} & 400 & 310  & 13.2 \\
            GMM SFS JM &    0.624 (0.028)             & 0.631 (0.034)             & 0.461 (0.027)             & 2   & 310  & 11 \\
            GMM SFFS JM &   0.624 (0.028)             & 0.631 (0.034)             & 0.461 (0.027)             & 2.6 & 310  & 11 \\
            GMM ridge &     0.632 (0.007)             & 0.592 (0.010)             & 0.433 (0.008)             & 10  & 2000 & all \\
            KNN &           0.637 (0.005)             & 0.607 (0.005)             & 0.478 (0.002)             &     &      & all \\
            Random Forest & {\bfseries 0.729 (0.004)} & {\bfseries 0.673 (0.005)} & {\bfseries 0.529 (0.009)} & 20  & 840  & all \\
            \bottomrule
        \end{tabular}
    \end{table}

    \begin{table}[!t]
        \centering
        \caption{Kappa coefficient and processing time for 50,000 samples by class and averaged over 5 trials (standard deviation in parenthesis). Processing times are given in second.\label{tab:potsdam-otbsimu-big}}
        \begin{tabular}{|lcccccc}\toprule
            & {\bfseries 5\_11 (train)} & {\bfseries 5\_12 (test)} & {\bfseries 3\_10 (test)} & {\bfseries Train. Time} & {\bfseries Classif. time} & {\bfseries \# of selected features} \\ \cline{2-6}
            GMM SFS kappa & 0.713 (0.001) & 0.684 (0.001) & 0.531 (0.005) & 20000 & 340 & 29 \\
            GMM SFS JM &    0.560 (0.111) & 0.576 (0.104) & 0.435 (0.085) & 6 & 330 & 10 \\
            GMM SFFS JM &   0.560 (0.111) & 0.576 (0.104) & 0.435 (0.085) & 6.6 & 340 & 10 \\
            GMM ridge &     0.641 (0.015) & 0.611 (0.026) & 0.440 (0.015) & 460 & 2000 & all \\
            Random Forest & {\bfseries 0.851 (0.001)} & {\bfseries 0.715 (0.001)} & {\bfseries 0.573 (0.002)} & 2000 & 2000 & all \\
            \bottomrule
        \end{tabular}
    \end{table}

>>>>

    The computation of~\ref{eq:decision} requires the inversion of the
    covariance matrix and and the  computation of the logarithm of the
    determinant.  The estimations of these terms suffer from the curse
    of  dimensionality\cite{bouveyron2014model}, because  in practice,
    the  number of  parameters  $\eta_c$ to  estimate  for each  class
    increases quadratically with respect to the number of features, as
    illustrated in Figure~\ref{fig:nb-param}.  Hence, if the number of
    observation  $n_c$ is  small  compared to  the  number of  samples
    $n_c$, the  estimated covariance  matrix is badly  conditioned and
    thus the computation  of its inverse and its  determinant would be
    unstable.   The worst  situation  is $n_c<\eta_c$  which leads  to
    singular covariance matrix.  Unfortunately, this situation happens
    regularly in remote sensing.   For instance in hyperspectral image
    classification,  very few  labeled samples  are usually  available
    because of the difficulty and the cost to collect ground-truth.

@@

Features selection algorithms can be divided into three types. The first type, called \emph{filter methods}, is based uniquely on data analysis. Features are ranked according to a statistical analysis of the data. For example, the Principal Component Analysis (PCA) described in \cite{jimenez1998supervised} is a typical filter method and there is numerous other methods \cite{bruzzone1995extension,biesiada2007feature,demir2008phase}.

The second sort are known as \emph{wrapper methods} which can be seen as search method to determine the best subset of variables for a given learning model. As exhaustive searches are to expansive in terms of processing time, several search strategies have been designed. Some are optimal under particular hypothesis \cite{narendra1977branch} and other sub-optimal but easier to set up \cite{whitney1971direct,somol1999adaptive}. The advantage of such methods compare to filter methods is that they are dedicated to a particular model but on the other hand, as they require the training of multiple models to test various set of variables, they tend to be more time consuming.

The third type corresponds to the \emph{embedded method} which do not separate the features selection process from the learning algorithm and allow interaction between the two processes. The basic example of such method is the decision tree algorithm in which a feature is selected for the creation of each node. Embedded methods also exist for other models, e.g. SVM \cite{guyon2002gene,weston2003use}.

The feature selection method proposed in this work is a wrapper method associated to GMM models. To set up a wrapper method, two elements are needed. A function that  ranks the features according to some criteria and a search strategy to optimize the function. The criterion used are described in Section~\ref{sec:criterion} and the two search strategy are discussed in~\ref{sec:selection:method}.

@@

        \subsection{Selection method}
        \label{sec:selection:method}

@@

\section{Datasets}
\label{sec:datasets}
Numerical experiments  have been conducted on  two different datasets.
The first  dataset, called  \emph{Aisa}, is an  airborne hyperspectral
dataset  and  the  second,  called \emph{Potsdam},  is  an  very  high
resolution multispectral airborne image.

@@

    \begin{table}[!t]
        \centering
        \caption{Kappa coefficient and processing time for 1,000 samples by class and averaged over 5 trials (standard deviation in parenthesis). Processing times are given in second.\label{tab:potsdam-otbsimu}}
        \begin{tabular}{lcccccc}\toprule
            & {\bfseries 5\_11 (train)} & {\bfseries 5\_12 (test)} & {\bfseries 3\_10 (test)} & {\bfseries Train. Time} & {\bfseries Classif. time} & {\bfseries \# of selected features} \\ \cline{2-7}
            GMM SFS kappa & 0.694 (0.002)             & {\bfseries 0.669 (0.005)} & {\bfseries 0.533 (0.008)} & 400 & 310  & 13.2 \\
            GMM SFS JM &    0.624 (0.028)             & 0.631 (0.034)             & 0.461 (0.027)             & 2   & 310  & 11 \\
            GMM SFFS JM &   0.624 (0.028)             & 0.631 (0.034)             & 0.461 (0.027)             & 2.6 & 310  & 11 \\
            GMM ridge &     0.632 (0.007)             & 0.592 (0.010)             & 0.433 (0.008)             & 10  & 2000 & all \\
            KNN &           0.637 (0.005)             & 0.607 (0.005)             & 0.478 (0.002)             &     &      & all \\
            Random Forest & {\bfseries 0.729 (0.004)} & {\bfseries 0.673 (0.005)} & {\bfseries 0.529 (0.009)} & 20  & 840  & all \\
            \bottomrule
        \end{tabular}
    \end{table}

    \begin{table}[!t]
        \centering
        \caption{Kappa coefficient and processing time for 50,000 samples by class and averaged over 5 trials (standard deviation in parenthesis). Processing times are given in second.\label{tab:potsdam-otbsimu-big}}
        \begin{tabular}{|lcccccc}\toprule
            & {\bfseries 5\_11 (train)} & {\bfseries 5\_12 (test)} & {\bfseries 3\_10 (test)} & {\bfseries Train. Time} & {\bfseries Classif. time} & {\bfseries \# of selected features} \\ \cline{2-7}
            GMM SFS kappa & 0.713 (0.001) & 0.684 (0.001) & 0.531 (0.005) & 20000 & 340 & 29 \\
            GMM SFS JM &    0.560 (0.111) & 0.576 (0.104) & 0.435 (0.085) & 6 & 330 & 10 \\
            GMM SFFS JM &   0.560 (0.111) & 0.576 (0.104) & 0.435 (0.085) & 6.6 & 340 & 10 \\
            GMM ridge &     0.641 (0.015) & 0.611 (0.026) & 0.440 (0.015) & 460 & 2000 & all \\
            Random Forest & {\bfseries 0.851 (0.001)} & {\bfseries 0.715 (0.001)} & {\bfseries 0.573 (0.002)} & 2000 & 2000 & all \\
            \bottomrule
        \end{tabular}
    \end{table}
